Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-05-25 manuscript/Manuscript_Claire.Rmd
Figures in text

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin&sect;
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage{microtype} 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and that many female researchers feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

@@

As expected, $\alpha'$ was usually almost identical to $\alpha$ (Figure S1), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (Figure S2). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation og $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore use the adjusted statistic $\alpha'$ in all analyses.

@@

@@

@@

@@

@@

\newpage
# Figures
![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations. Specifically, there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers we would expect under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals. The Wahlund effect will tend to inflate the frequency of same-sex coauthors whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

\newpage
![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after adjusting p-values using Benjamini-Hochberg false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/figure2.pdf)

\newpage
![There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x-axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3. \label{alpha_gender_ratio}](../figures/figure3.pdf)

\newpage
![Journal impact factor (expressed relative to the average for the discipline) is negatively correlated with $\alpha'$. The relationship is noisy ($R^2$ = `r R2`), but the results suggest that journals with strong homophily tend to have lower impact factors than journals with weak homophily in the same discipline. \label{impact_factor}](../figures/figure4.pdf)

\newpage
![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. \label{simulation}](../figures/figure5.pdf)

>>>>

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin&sect;
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage{microtype} 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor, as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations. Specifically, there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers we would expect under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals. The Wahlund effect will tend to inflate the frequency of same-sex coauthors whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

@@

As expected, $\alpha'$ was usually almost identical to $\alpha$ (Figure S1), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (Figure S2). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation of $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore use the adjusted statistic $\alpha'$ in all analyses.

@@

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after adjusting p-values using Benjamini-Hochberg false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/figure2.pdf)

@@

![There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x-axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3. \label{alpha_gender_ratio}](../figures/figure3.pdf)

@@

![Journal impact factor (expressed relative to the average for the discipline) is negatively correlated with $\alpha'$. The relationship is noisy ($R^2$ = `r R2`), but the results suggest that journals with strong homophily tend to have lower impact factors than journals with weak homophily in the same discipline. \label{impact_factor}](../figures/figure4.pdf)

@@

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. \label{simulation}](../figures/figure5.pdf)

@@
